---
title: Visualizing COVID-19 Data Using R
author: Sharan S.
output: html_document
---

Researchers at [Our World in Data](https://ourworldindata.org/coronavirus-source-data) have aggregated data on COVID-19 cases, deaths, vaccinations, etc from various sources such as WHO and Johns Hopkins University, and made it available in a GitHub repository.

I will be using this dataset to visualize various COVID-19 related metrics, and also test some hypotheses relating to the nature of COVID-19 cases.

```{r, include=FALSE}
library(skimr)
library(tidyverse)
```

### Total Per-Capita COVID-19 cases in Canada over time

First, I'll import the dataset as such.

```{r, message=FALSE, cache=TRUE}
canada_cases <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/jhu/total_cases_per_million.csv")
```

The raw data has each country's cases in its own column, and dates in rows. To tidy this data, I'll first filter out only the relevant countries and gather the country info under one column.

```{r}
canada_cases <- canada_cases %>%
  select(date, World, Canada, `United States`) %>%
  gather(key = country, value = cases, -date)

skim_without_charts(canada_cases)
```

Now, this data can be visualized using `ggplot2`.

```{r, fig.align="center"}
ggplot(canada_cases, aes(x = date, y = cases)) +
  geom_line(aes(color = country)) +
  scale_y_log10() +
  scale_x_date(date_labels = "%b %y", date_breaks = "2 months",
    guide = guide_axis(angle = 60)) +
  labs(title = "Total Per-Capita COVID-19 Cases Over Time",
    x = "Time", y = "Confirmed COVID-19 cases (per million)",
    caption = "Source: Our World in Data")
```
